Coherence between subjective feeling and facial expression: A comparison of human and automatic recognition

### 2019-04-18 ACII_2019_paper/ACII_2019_word.Rmd
word

---
title: "Coherence between subjective feeling and facial expression: a comparison of human and automatic recognition."
output: word_document
bibliography: mybibfile.bib
abstract: |
  While it has been taken for granted in the development of several automatic facial expression recognition tools, the question of the link between subjective feelings and facial expressions is still a subject of debate. On one hand the behaviorist approach conceives emotions as genetically hardwired and therefore being genuinely displayed through facial expressions. On the other hand the constructivist approach conceives emotions as socially constructed, the emotional meaning of a facial expression being inferred by the observer. In order to evaluate the link between the subjective feeling of emotions and their recognition based on facial expression, 232 videos of encoders recruited to carry out an emotion elicitation task were annotated by 1383 human observers as well as by an automatic facial expression classifier. Results show a low accuracy of human observers and of the automatic classifier to infer the subjective feeling from the facial expressions displayed by encoders . They also show a weak link between self-reported emotional states and facial emotional displays. Based on these results, the hypothesis of genetically hardwired emotion genuinely display is difficult to support whereas the idea of emotion and facial expression socially constructed appears to be more likely. Then automatic emotion recognition tools  based on facial expressions should be questioned.

---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  cache.extra = knitr::rand_seed, 
  message = FALSE, 
  warning = FALSE, 
  error = FALSE, 
  echo = FALSE,
  fig.pos='H')
#
set.seed(123)# Seed for random number generation
options(scipen = 999)# Disable scientific number format
#
library(kableExtra)
library(tidyverse) # Data wrangling & grammar of graphics
library(here) # Locate files relative to project root
library(caret)
library(plotROC)
library(pROC)
library(pander)
library(papaja)
library(captioner)
table_nums <- captioner::captioner(prefix = "Table")
table_nums(
  name = "confusionTable_sr_hr", 
  caption = "Agreement accuracy metrics for each emotion.")
table_nums(
  name = "confusionTable_sr_ar", 
  caption = "Agreement accuracy metrics for each emotion.")
#
fig_nums <- captioner::captioner()
fig_nums(
  name = "dynemo_img", 
  caption = "Example of a front facing recording synced with the full view of the participant and the elicitation task. This picture is taken from a pilot with projects collaborators and all gave a consent for the publication of their photos and videos.")
fig_nums(
  name = "confusionMatrix_sr_hr", 
  caption = "Confusion matrix of between the emotion self-reported as being characteristic of the elicitation with the emotion recognized by the human observers.")
fig_nums(
  name = "confusionMatrix_sr_ar", 
  caption = "Confusion matrix of between the emotion self-reported as being characteristic of the elicitation with the emotion recognized by the automatic classifier.")
fig_nums(
  name = "confusionMatrix_hr_ar", 
  caption = "Confusion matrix of between the human obersers and automatic classifier recognition.")
```

@@

Introduction
=============

With the development of commercial automatic facial expression recognition tools [see @dupre2018accuracy for a non-exhaustive list of available tools]), industries and governments are gradually implementing this technology in order to track humans’ emotions in various scenarios (e.g., marketing, healthcare, automotive to name a few). This technoloogy rests on the premise that facial expressions provide a direct access to individuals’ subjective feeling. Even if this premise is central to the modern mainstream approach of human emotion, recent research in affective science are challenging it. Once the two competing approaches briefly described as well as the predictions they respectively entail, an experiment testing these hypotheses will be presented and its results analyzed in order to provide empirical evidence to contribute to answer the question.

## The link between emotion and facial expression through the behaviorist approach
Based on the behaviorist approach initiated by Darwin in *The Expression of the Emotions in Man and Animals* [@darwin1872expression], facial expressions are conceived as a genuine displays of individuals’ inner emotional state. This hypothesis is used as a basis for the Basic Emotion Theory (BET) which states that a set of six emotions are universally displayed and are genetically hardwired not only in humans [@ekman1992argument] but also in different animal species [@de2019mama]. According to this view, “when emotions are aroused by perception of a social event, a set of central commands produce patterned emotion-specific changes in multiple systems, including […] facial expressions.” [@ekman2007directed, p49]). To cope with critics, several amendments have been made to the BET, increasing the number of basic emotions from six to seven [@ekman1988universality] ) as well as adding the concept of “display rules” to explain cultural differences in the management of facial expressions [@ekman1987universals].

@@

## The link between emotion and facial expression through the social constructivist approach
Detractors of the Basic Emotion Theory consider emotion not as genetically hardwired but as a learnt association between a given situation and an appropriate response [@averill1980constructivist; @barrett2017emotions]). For the tenants of the constructivist approach, emotions are “concepts” based on past experiences and which are “a collection of embodied, whole brain representations that predict what is about to happen in the sensory environment, what the best action is to deal with impending events, and their consequences for allostasis” [@barrett2017theory, p12]. Following this assumption, faces are best conceived as tools displaying signals in social interactions [@crivelli2018facial]. These signals can convey individuals’ motivations and readiness [@frijda1997facial] or social messages [@fridlund1995human]. Therefore, facial expressions are thought as behaviors which meaning is inferred by the observer. Findings support this observer dependence [@lindquist2013s; @niedenthal2017embodied]. They show that to make meaning of another person’s facial behavior, the perceiver relies in particular on her/his knowledge about emotion categories.

## The recognition of emotional facial expressions
Regarding emotional facial expression (EFE) recognition, the behaviorist approach assertions and the constructivist approach claims lead to two divergent predictions. The foremost postulates that, when triggered, each basic emotion is expressed by a prototypical face (non basic emotions being blends of the basic ones). Basic emotions are easily recognized by all human observers and emotional states are accessible by facial measurement. As the recognition of facial expressions is based on the identification of specific patterns of facial movements, it implies that EFE recognition of both human observers and automatic classifiers should be as accurate. The constructivist approach, in contrast, affirms that facial expressions do not provide a direct access to individuals’ subjective feeling. Even if the face possibly moves during an emotional episode, facial muscle movements are not linked in a one-to-one manner to a specific discrete emotional experience [@doyle2017language, p418]. Instead, emotions are mentally constructed by the perceiver and mental categories of emotions are needed to accurately categorize facial movements. Therefore, since there is no emotional prototypical face, one should expect human observers’ superior ability to accurately recognize EFE as compared to automatic EFE recognition tools.

@@

For the emotion elicitation experiment, `r nrow(self_report_data)` French participants (`r self_report_gender[self_report_gender$genre_c == "F","n"]` females, `r self_report_gender[self_report_gender$genre_c == "H","n"]` males, *M*age = `r self_report_age[,"Age_m"]`, *SD*age = `r self_report_age[,"Age_sd"]`) were recruited to perform one out of 11 emotion elicitation tasks designed to trigger a positive, a specific negative or a neutral emotional state. Encoders’ face were recorded using an hidden camera resulting `r nrow(self_report_data)` front facing 768x576 videos varying from `r min(metadata_video$ffprobe_duration)`s to `r max(metadata_video$ffprobe_duration)`s (`r fig_nums("dynemo_img", display = "cite")`). These recordings form the DynEmo database [see @tcherkassof2013dynemo for a full description of tasks and procedure].

```{r dynemo_img, fig.cap=fig_nums("dynemo_img")}
knitr::include_graphics(here::here("./img/dynemo_img.jpg"))
```

After the emotion elicitation task the encoders rated their subjective feeling on Likert scales from 0 ("not at all") to 5 ("strongly") related to six “basic” emotion labels (i.e., *anger*, *disgust*, *fear*, *happiness*, *surprise* and *sadness*) as well as six “non-basic” emotion labels (i.e., *pride*, *curiosity*, *boredom*, *shame*, *humiliation*, and *disappointment*).

Finally, a debriefing session was performed to ensure that encoders were not durably affected by the emotion elicitation task. The debriefing was also used to check that encoders did not guess the real purpose of the experiment (e.g. being filmed while they were performing an emotional elicitation task) to guarantee facial expressions’ genuineness. All encoders gave their agreement on their data and video to be processed for research purpose only.

## Human Facial Expression Recognition

@@

For the human facial expression recognition method, `r nrow(unique_ppt_count)` student participants were recruited to annotate `r nrow(unique_video_count)` out of the `r nrow(self_report_data)` video, therefore only the `r nrow(unique_video_count)` annotated videos will be analysed in this paper. Because videos have different durations, participants had to annotate a series of video corresponding to 30min long in total. Each video was annotated `r mean(unique_video_count$n) %>% round(0)` times on average (*SD* = `r sd(unique_video_count$n) %>% round(0)`).

The annotation of facial expressions was performed on-site using *Oudjat*, a software for designing video annotation experiments [@dupre2015oudjat]. For each video, the annotation procedure followed two steps. First, the participants had to identify the emotional sequences by pressing the space bar of their keyboard to indicate the beginning and the end of the emotional sequences while watching the video. Second, the participants watched each emotional sequence previously identified and labeled the sequence using one of the 12 emotions proposed including six “basic” emotion labels (i.e., *anger*, *disgust*, *fear*, *happiness*, *surprise* and *sadness*) and six "non-basic" emotion labels (i.e., *pride*, *curiosity*, *boredom*, *shame*, *humiliation*, and *disappointment*). They also had the possibility to indicate that the sequence was expressing none of the proposed emotion labels.

This annotation procedure results in a uni-dimensional time-series for each video per human observer identifying for each second of the video which emotion was recognized. Then, time-series corresponding to the same video were aggregated to calculate the proportion of human observers $x_{video_{i}.label_{j}.t_{k}}$ for each second of the video per emotional label \ref{eq:1}.

@@

## Automatic Facial Expression Recognition

The `r nrow(unique_video_count)` annotated video were processed with Affdex (SDK v3.4.1). Affdex is an automatic facial expression recognition classifier developed and distributed by Affectiva is a spin-off company resulting from the research activities of MIT media lab created in 2009 [@mcduff2016affdex]. Affdex’s algorithm uses Histogram of Oriented Gradient (HOG) features and Support Vector Machine (SVM) classifiers in order to recognize facial expressions. For each video frame, Affdex identifies the probability $p_{video_{i}.label_{j}.t_{k}}$ from 0 to 100 (rescaled to 0 to 1 for the analysis) of the face as expressing one of the six “basic” emotion labels (i.e., *anger*, *disgust*, *fear*, *happiness*, *surprise* and *sadness*) as well as additional psychological states such as *valence*, *engagement* or *contempt*, and facial features such as *cheek raise*, *eye widen* or *jaw drop*.

For both human and automatic recognition, to determine which of the six “basic” emotions can be used to identify each video, the recognition probability for each label by frame was converted into odd ratio by label [@dente2017measures]. The highest sum of each odd ratio time-series defines the label recognized by the automatic classifier (\ref{eq:1} for human recognition and \ref{eq:1} for automatic recognition).
\begin{equation}
 \label{eq:2}
video_{i}.label = \max\left(\frac{\sum_{k=1}^{n}x_{video_{i}.label_{j}.t_{k}}}{\sum_{k=1}^{n}x_{video_{i}.t_{k}}}\right)
\end{equation}

@@

Encoders’ subjective feeling is compared with human observers recognition in a confusion matrix (`r fig_nums("confusionMatrix_sr_hr", display = "cite")`). Each emotion label used to describe encoders’ self-reported subjective feeling (i.e., the label rated with the highest value) is compared with the emotion labels which were rated with the highest score by human observers.

```{r confusionMatrix_sr_hr, fig.height=4, fig.cap=fig_nums("confusionMatrix_sr_hr")}
confusionMatrix_sr_hr_freq <- confusionMatrix_sr_hr$table %>% 
  as.data.frame() %>% 
  dplyr::rename(hr_emotion = Prediction, sr_emotion = Reference) %>% 
  dplyr::mutate(n_video = sum(Freq)) %>% 
  dplyr::mutate(prop_video = Freq/sum(Freq)) %>% 
  dplyr::group_by(sr_emotion) %>% 
  dplyr::mutate(n_sr = sum(Freq)) %>%
  dplyr::mutate(prop_sr = Freq/sum(Freq)) %>% 
  dplyr::ungroup()

@@

Results of the confusion matrix show a moderate agreement between the emotion felt by the encoder during the elicitation and the emotion recognized by the human observers (Accuracy = `r round(confusionMatrix_sr_hr$overall["Accuracy"],2)`, 95%CI[`r round(confusionMatrix_sr_hr$overall["AccuracyLower"],2)`,`r round(confusionMatrix_sr_hr$overall["AccuracyUpper"],2)`]; Kappa = `r round(confusionMatrix_sr_hr$overall["Kappa"],2)`) except for *disgust* (`r scales::percent(as.numeric(confusionMatrix_sr_hr_freq[confusionMatrix_sr_hr_freq$hr_emotion == "disgust" & confusionMatrix_sr_hr_freq$sr_emotion == "disgust", "prop_sr"]))` of the videos self-reported). Sensitivity, specificity, precision and F1 score for each emotion reveals that *happiness* has the highest coherence ratio whereas *sadness* has the lowest coherence ratio between true positives and false positives `r table_nums("confusionTable_sr_hr", display = "cite")`. Interestingly human observers seem to recognize *surprise* expressed in videos where *anger*, *fear* *happiness* and *sadness* was the highest self-reported emotion (respectively `r scales::percent(as.numeric(confusionMatrix_sr_hr_freq[confusionMatrix_sr_hr_freq$hr_emotion == "surprise" & confusionMatrix_sr_hr_freq$sr_emotion == "anger", "prop_sr"]))`, `r scales::percent(as.numeric(confusionMatrix_sr_hr_freq[confusionMatrix_sr_hr_freq$hr_emotion == "surprise" & confusionMatrix_sr_hr_freq$sr_emotion == "fear", "prop_sr"]))`, `r scales::percent(as.numeric(confusionMatrix_sr_hr_freq[confusionMatrix_sr_hr_freq$hr_emotion == "surprise" & confusionMatrix_sr_hr_freq$sr_emotion == "happiness", "prop_sr"]))` and `r scales::percent(as.numeric(confusionMatrix_sr_hr_freq[confusionMatrix_sr_hr_freq$hr_emotion == "surprise" & confusionMatrix_sr_hr_freq$sr_emotion == "sadness", "prop_sr"]))` of the videos self-reported), and in a lower instance *happiness* was recognized in videos where *fear* and *surprise* was the highest self-reported emotion (respectively `r scales::percent(as.numeric(confusionMatrix_sr_hr_freq[confusionMatrix_sr_hr_freq$hr_emotion == "happiness" & confusionMatrix_sr_hr_freq$sr_emotion == "fear", "prop_sr"]))` and `r scales::percent(as.numeric(confusionMatrix_sr_hr_freq[confusionMatrix_sr_hr_freq$hr_emotion == "happiness" & confusionMatrix_sr_hr_freq$sr_emotion == "surprise", "prop_sr"]))` of the videos self-reported).

@@

# confusionMatrix_sr_hr_table %>%
#   knitr::kable(
#     "latex",
#     caption = "Agreement accuracy metrics for each emotion. ",
#     booktabs = TRUE,
#     digits = 2,
#     linesep = "",
#     escape = FALSE,
#     align = c('l',rep('c', 4))
#   ) %>%
#   kableExtra::kable_styling(
#     font_size = 8,
#     full_width = TRUE,
#     latex_options = c("hold_position")) %>%
#    add_footnote("Note. \\textit{na.} values are produced when not enough data are available to compute accuracy indicators.", notation="none", escape = FALSE)

confusionMatrix_sr_hr_table %>%
  papaja::apa_table(
    caption = table_nums("confusionTable_sr_hr"),
    row.names = FALSE,
    format = "word",
    escape = FALSE
    )
```

However, self-reports show a very high proportion  of undetermined emotional states which reveals the potential limit of using 6-points likert scales to measure emotional self-reports. Indeed, encoders can easily score to the maximum for more than one emotion.

## Correlation between self-report and automatic facial expression recognition

@@

Similarly to the previous analysis, a confusion matrix was used to compare encoders’ subjective feeling with the emotion label recognized by the automatic classifier (`r fig_nums("confusionMatrix_sr_ar", display = "cite")`).

```{r confusionMatrix_sr_ar, fig.height=4, fig.cap=fig_nums("confusionMatrix_sr_ar")}
confusionMatrix_sr_ar_freq <- confusionMatrix_sr_ar$table %>% 
  as.data.frame() %>% 
  dplyr::rename(ar_emotion = Prediction, sr_emotion = Reference) %>% 
  dplyr::mutate(n_video = sum(Freq)) %>% 
  dplyr::mutate(prop_video = Freq/sum(Freq)) %>% 
  dplyr::group_by(sr_emotion) %>% 
  dplyr::mutate(n_sr = sum(Freq)) %>%
  dplyr::mutate(prop_sr = Freq/sum(Freq)) %>% 
  dplyr::ungroup()

@@

# confusionMatrix_sr_ar_table %>%
#   knitr::kable(
#     "latex",
#     caption = "Agreement accuracy metrics for each emotion.",
#     booktabs = T,
#     digits = 2,
#     linesep = "",
#     escape = FALSE,
#     align = c('l',rep('c', 4))
#   ) %>%
#   kableExtra::kable_styling(
#     font_size = 8,
#     full_width = TRUE,
#     latex_options = c("hold_position")) %>%
#    add_footnote("Note. \\textit{na.} values are produced when not enough data are available to compute accuracy indicators.", notation="none", escape = FALSE)

>>>>

---
title: "Coherence between subjective feeling and facial expression: A comparison of human and automatic recognition" 
affiliation:
  institution-columnar: true

  institution:
    - name: Dublin City University
      location: Dublin, Ireland
      email: damien.dupre@dcu.ie
      mark: 1
      author:
        - name: Damien Dupré
    - name: Grenoble Alpes University
      location: Grenoble, France
      mark: 2
      author:
        - name: Anna Tcherkassof
          email: anna.tcherkassof@univ-grenoble-alpes.fr
abstract: |
  While it has been taken for granted in the development of several automatic facial expression recognition tools, the question of the link between subjective feelings and facial expressions is still a subject of debate. On one hand the behaviorist approach conceives emotions as genetically hardwired and therefore being genuinely displayed through facial expressions. On the other hand the constructivist approach conceives emotions as socially constructed, the emotional meaning of a facial expression being inferred by the observer. In order to evaluate the link between the subjective feeling of emotions and their recognition based on facial expression, 232 videos of encoders recruited to carry out an emotion elicitation task were annotated by 1383 human observers as well as by an automatic facial expression classifier. Results show a low accuracy of human observers and of the automatic classifier to infer the subjective feeling from the facial expressions displayed by encoders . They also show a weak link between self-reported emotional states and facial emotional displays. Based on these results, the hypothesis of genetically hardwired emotion genuinely display is difficult to support whereas the idea of emotion and facial expression socially constructed appears to be more likely. Then automatic emotion recognition tools based on facial expressions should be questioned.

header-includes:
  - \usepackage{booktabs}
  - \usepackage{float}
  - \usepackage{tabu}
  - \usepackage{wrapfig}
  - \usepackage[none]{hyphenat}
  - \floatplacement{figure}{H}

bibliography: mybibfile.bib
output: word_document
csl: http://www.zotero.org/styles/ieee
fig_caption: yes
keep_tex: yes
number_sections: yes
includes:
  in_header: preamble-latex.tex
---

\begin{IEEEkeywords}
Facial expression, self-report, human observer, automatic recognition.
\end{IEEEkeywords}

```{r setup, include = FALSE}
# options ----------------------------------------------------------------------
knitr::opts_chunk$set(
  cache.extra = knitr::rand_seed, 
  message = FALSE, 
  warning = FALSE, 
  error = FALSE, 
  echo = FALSE)

set.seed(123)# Seed for random number generation
options(scipen = 999)# Disable scientific number format

# packages ----------------------------------------------------------------------
library(kableExtra) # Tables
library(tidyverse) # Data wrangling & grammar of graphics
library(here) # Locate files relative to project root
library(caret) # Confusion matrices
library(papaja) # Correlation results
```

@@

I. Introduction
=============

With the development of commercial automatic facial expression recognition tools [see @dupre2018accuracy for a non-exhaustive list of available tools], industries and governments are gradually implementing this technology in order to track humans’ emotions in various scenarios (*e.g.*, marketing, healthcare, automotive to name a few). This technoloogy rests on the premise that facial expressions provide a direct access to individuals’ subjective feeling. Even if this premise is central to the modern mainstream approach of human emotion, recent research in affective science are challenging it. Once the two competing approaches briefly described as well as the predictions they respectively entail, an experiment testing these hypotheses will be presented and its results analyzed in order to provide empirical evidence to contribute to answer the question.

## A. The Behaviorist Approach
Based on the behaviorist approach initiated by Darwin in *The Expression of the Emotions in Man and Animals* [@darwin1872expression], facial expressions are conceived as a genuine displays of individuals’ inner emotional state. This hypothesis is used as a basis for the Basic Emotion Theory (BET) which states that a set of six emotions are universally displayed and are genetically hardwired not only in humans [@ekman1992argument] but also in different animal species [@de2019mama]. According to this view, *“when emotions are aroused by perception of a social event, a set of central commands produce patterned emotion-specific changes in multiple systems, including […] facial expressions.”* [@ekman2007directed, 49]. To cope with critics, several amendments have been made to the BET, increasing the number of basic emotions from six to seven [@ekman1988universality] as well as adding the concept of “display rules” to explain cultural differences in the management of facial expressions [@ekman1987universals].

@@

## B. The Social Constructivist Approach
Detractors of the Basic Emotion Theory consider emotion not as genetically hardwired but as a learnt association between a given situation and an appropriate response [@averill1980constructivist; @barrett2017emotions]. For the tenants of the constructivist approach, emotions are “concepts” based on past experiences and which are *“a collection of embodied, whole brain representations that predict what is about to happen in the sensory environment, what the best action is to deal with impending events, and their consequences for allostasis”* [@barrett2017theory, 12]. Following this assumption, faces are best conceived as tools displaying signals in social interactions [@crivelli2018facial]. These signals can convey individuals’ motivations and readiness [@frijda1997facial] or social messages [@fridlund1994human]. Therefore, facial expressions are thought as behaviors which meaning is inferred by the observer. Findings support this observer dependence [@lindquist2013s; @niedenthal2017embodied]. They show that to make meaning of another person’s facial behavior, the perceiver relies in particular on her/his knowledge about emotion categories.

## C. Emotional Facial Expression Recognition
Regarding emotional facial expression (EFE) recognition, the behaviorist approach assertions and the constructivist approach claims lead to two divergent predictions. The foremost postulates that, when triggered, each basic emotion is expressed by a prototypical face (non basic emotions being blends of the basic ones). Basic emotions are easily recognized by all human observers and emotional states are accessible by facial measurement. As the recognition of facial expressions is based on the identification of specific patterns of facial movements, it implies that EFE recognition of both human observers and automatic classifiers should be as accurate. The constructivist approach, in contrast, affirms that facial expressions do not provide a direct access to individuals’ subjective feeling. Even if the face possibly moves during an emotional episode, facial muscle movements are not linked in a one-to-one manner to a specific discrete emotional experience [@doyle2017language, p418]. Instead, emotions are mentally constructed by the perceiver and mental categories of emotions are needed to accurately categorize facial movements. Therefore, since there is no emotional prototypical face, one should expect human observers’ superior ability to accurately recognize EFE as compared to automatic EFE recognition tools (Figure \ref{fig:models_img}).

```{r models_img, fig.height=5, fig.cap="\\label{fig:models_img}Comparison of behaviorist and social constructivist approaches according to the expected differences in emotion recognition."}
# knitr::include_graphics("C:/Users/dupred/Desktop/Projects/dynemo_feel_express/img/dynemo_img.png", auto_pdf = FALSE)
raster::stack(here::here("./img/models.png")) %>%
  raster::plotRGB()
```

@@

For the emotion elicitation experiment, `r nrow(self_report_data)` French participants (`r self_report_gender[self_report_gender$genre_c == "F","n"]` females, `r self_report_gender[self_report_gender$genre_c == "H","n"]` males, *M*~age~ = `r self_report_age[,"Age_m"]`, *SD*~age~ = `r self_report_age[,"Age_sd"]`) were recruited to perform one out of 11 emotion elicitation tasks designed to trigger a positive, a specific negative or a neutral emotional state. Encoders’ face were recorded using an hidden camera resulting `r nrow(self_report_data)` front facing 768x576 videos varying from `r min(metadata_video$ffprobe_duration)`s to `r max(metadata_video$ffprobe_duration)`s (Figure \ref{fig:dynemo_img}). These recordings form the DynEmo database [see @tcherkassof2013dynemo for a full description of tasks and procedure].

```{r dynemo_img, fig.height=3, fig.cap="\\label{fig:dynemo_img}Example of a front facing recording synced with the full view of the participant and the elicitation task. This picture is taken from a pilot with projects collaborators and all gave a consent for the publication of their photos and videos."}
raster::stack(here::here("./img/dynemo_img.png")) %>%
  raster::plotRGB()
```

After the emotion elicitation task the encoders rated their subjective feeling on Likert scales from 0 ("not at all") to \nolinebreak 5 \nolinebreak ("strongly") related to six “basic” emotion labels (*i.e.*, *anger*, *disgust*, *fear*, *happiness*, *surprise* and *sadness*) as well as six “non-basic” emotion labels (*i.e.*, *pride*, *curiosity*, *boredom*, *shame*, *humiliation*, and *disappointment*).

Finally, a debriefing session was performed to ensure that encoders were not durably affected by the emotion elicitation task. The debriefing was also used to check that encoders did not guess the real purpose of the experiment (*e.g.*, being filmed while they were performing an emotional elicitation task) to guarantee facial expressions’ genuineness. All encoders gave their agreement on their data and video to be processed for research purpose only.

## B. Human Facial Expression Recognition

@@

For the human facial expression recognition method, `r nrow(unique_ppt_count)` \nolinebreak student participants were recruited to annotate `r nrow(unique_video_count)` out of the `r nrow(self_report_data)` videos, therefore only the `r nrow(unique_video_count)` annotated videos will be analysed in this paper. Because videos have different durations, participants had to annotate a series of video corresponding to 30min long in total. Each video was annotated `r mean(unique_video_count$n) %>% round(0)` times on average (*SD* = `r sd(unique_video_count$n) %>% round(0)`).

The annotation of facial expressions was performed on-site using *Oudjat*, a software for designing video annotation experiments [@dupre2015oudjat]. For each video, the annotation procedure followed two steps. First, the participants had to identify the emotional sequences by pressing the space bar of their keyboard to indicate the beginning and the end of the emotional sequences while watching the video. Second, the participants watched each emotional sequence previously identified and labeled the sequence using one of the 12 \nolinebreak emotions proposed including six “basic” emotion labels (*i.e.*, *anger*, *disgust*, *fear*, *happiness*, *surprise* and *sadness*) and six "non-basic" emotion labels (*i.e.*, *pride*, *curiosity*, *boredom*, *shame*, *humiliation*, and *disappointment*). They also had the possibility to indicate that the sequence was expressing none of the proposed emotion labels.

This annotation procedure results in a uni-dimensional time-series for each video per human observer identifying for each second of the video which emotion was recognized. Then, time-series corresponding to the same video were aggregated to calculate the proportion of human observers $x_{video_{i}.label_{j}.t_{k}}$ for each second of the video per emotional label (EQ\ref{eq:1}).

@@

## C. Automatic Facial Expression Recognition

The `r nrow(unique_video_count)` annotated video were processed with Affdex (SDK v3.4.1). Affdex is an automatic facial expression recognition classifier developed and distributed by Affectiva is a spin-off company resulting from the research activities of MIT media lab created in 2009 [@mcduff2016affdex]. Affdex’s algorithm uses Histogram of Oriented Gradient (HOG) features and Support Vector Machine (SVM) classifiers in order to recognize facial expressions. For each video frame, Affdex identifies the probability $p_{video_{i}.label_{j}.t_{k}}$ from 0 to 100 (rescaled to 0 to 1 for the analysis) of the face as expressing one of the six “basic” emotion labels (*i.e.*, *anger*, *disgust*, *fear*, *happiness*, *surprise* and *sadness*) as well as additional psychological states such as *valence*, *engagement* or *contempt*, and facial features such as *cheek raise*, *eye widen* or *jaw drop*.

For both human and automatic recognition, to determine which of the six “basic” emotions can be used to identify each video, the recognition probability for each label by frame was converted into odd ratio by label [@dente2017measures]. The highest sum of each odd ratio time-series defines the label recognized by the automatic classifier (EQ\ref{eq:2} for human recognition and EQ\ref{eq:3} for automatic recognition).
\begin{equation}
\label{eq:2}
video_{i}.label = \max\left(\frac{\sum_{k=1}^{n}x_{video_{i}.label_{j}.t_{k}}}{\sum_{k=1}^{n}x_{video_{i}.t_{k}}}\right)
\end{equation}

@@

The overall correlation of recognition and non-recognition between self-reported emotions and human observers recognition is significant but low (`r corr_sr_hr$full_result`). In order to identify differences according the emotional labels, encoders’ subjective feeling is compared with human observers recognition in a confusion matrix (Figure \ref{fig:confusionMatrix_sr_hr}).

```{r confusionMatrix_sr_hr, fig.height=6, fig.cap="\\label{fig:confusionMatrix_sr_hr}Confusion matrix of between the emotion self-reported as being characteristic of the elicitation with the emotion recognized by the human observers."}
confusionMatrix_sr_hr_freq <- confusionMatrix_sr_hr$table %>% 
  as.data.frame() %>% 
  dplyr::rename(hr_emotion = Prediction, sr_emotion = Reference) %>% 
  dplyr::mutate(n_video = sum(Freq)) %>% 
  dplyr::mutate(prop_video = Freq/sum(Freq)) %>% 
  dplyr::group_by(sr_emotion) %>% 
  dplyr::mutate(n_sr = sum(Freq)) %>%
  dplyr::mutate(prop_sr = Freq/sum(Freq)) %>% 
  dplyr::ungroup()

@@

Each emotion label used to describe encoders’ self-reported subjective feeling (*i.e.*, the label rated with the highest value) is compared with the emotion labels which were rated with the highest score by human observers. Results of the confusion matrix show a moderate agreement between the emotion felt by the encoder during the elicitation and the emotion recognized by the human observers (Accuracy \nolinebreak = \nolinebreak `r round(confusionMatrix_sr_hr$overall["Accuracy"],2)`, 95% CI [`r round(confusionMatrix_sr_hr$overall["AccuracyLower"],2)`,`r round(confusionMatrix_sr_hr$overall["AccuracyUpper"],2)`]; Kappa = `r round(confusionMatrix_sr_hr$overall["Kappa"],2)`) except for *disgust* (`r scales::percent(as.numeric(confusionMatrix_sr_hr_freq[confusionMatrix_sr_hr_freq$hr_emotion == "disgust" & confusionMatrix_sr_hr_freq$sr_emotion == "disgust", "prop_sr"]))` of the videos self-reported). These results are far from those classicaly obtained in the literature for emotional facial expression recognition which ranges between 60% and 80% accuracy. However these results are mostly obtained with static (*i.e.,* pictures) and posed (*i.e.,* displayed by actors) facial expressions using only 6 emotional labels in a forced-choice paradigm.

Interestingly human observers seem to recognize *surprise* expressed in videos where *anger*, *fear* *happiness* and *sadness* was the highest self-reported emotion (respectively `r scales::percent(as.numeric(confusionMatrix_sr_hr_freq[confusionMatrix_sr_hr_freq$hr_emotion == "surprise" & confusionMatrix_sr_hr_freq$sr_emotion == "anger", "prop_sr"]))`, `r scales::percent(as.numeric(confusionMatrix_sr_hr_freq[confusionMatrix_sr_hr_freq$hr_emotion == "surprise" & confusionMatrix_sr_hr_freq$sr_emotion == "fear", "prop_sr"]))`, `r scales::percent(as.numeric(confusionMatrix_sr_hr_freq[confusionMatrix_sr_hr_freq$hr_emotion == "surprise" & confusionMatrix_sr_hr_freq$sr_emotion == "happiness", "prop_sr"]))` and `r scales::percent(as.numeric(confusionMatrix_sr_hr_freq[confusionMatrix_sr_hr_freq$hr_emotion == "surprise" & confusionMatrix_sr_hr_freq$sr_emotion == "sadness", "prop_sr"]))` of the videos self-reported), and in a lower instance *happiness* was recognized in videos where *fear* and *surprise* was the highest self-reported emotion (respectively `r scales::percent(as.numeric(confusionMatrix_sr_hr_freq[confusionMatrix_sr_hr_freq$hr_emotion == "happiness" & confusionMatrix_sr_hr_freq$sr_emotion == "fear", "prop_sr"]))` and `r scales::percent(as.numeric(confusionMatrix_sr_hr_freq[confusionMatrix_sr_hr_freq$hr_emotion == "happiness" & confusionMatrix_sr_hr_freq$sr_emotion == "surprise", "prop_sr"]))` of the videos self-reported).

Sensitivity, specificity, precision and F1 score for each emotion reveals that *happiness* has the highest coherence ratio whereas *sadness* has the lowest coherence ratio between true positives and false positives (Table \ref{table:confusionTable_sr_hr}).

@@

confusionMatrix_sr_hr_table %>%
  knitr::kable(
    "latex",
    caption = "\\label{table:confusionTable_sr_hr}Human recognition accuracy metrics for each emotion.",
    booktabs = TRUE,
    digits = 2,
    linesep = "",
    escape = FALSE,
    align = c('l',rep('c', 4))
  ) %>%
  kableExtra::kable_styling(
    latex_options = "HOLD_position",
    font_size = 8,
    full_width = TRUE 
    ) %>%
   add_footnote("Note. \\textit{na.} values are produced when not enough data are available to compute accuracy indicators.", notation="none", escape = FALSE)
```

Accuracy metrics by emotional labels indicate a discrepancy in the ratio of true/false positives. Whereas *happiness* and *disgust* obtain the highest scores, *anger*, *surprise* and *sadness* have the lowest recognition ratio. The underlying effect of expression intensity may explain why *happiness* and *disgust* are easily recognized. *Anger* and *sadness* as non-socialy desirable emotion may be have been felt but not expressed.

However, self-reports show a significant proportion of undetermined emotional states (`r scales::percent(nrow(list_automatic_ties)/358)` of the 358 videos) which reveals the potential limit of using 6-points likert scales to measure emotional self-reports. Indeed, encoders can easily score to the maximum for more than one emotion.

## B. Automatic Classifier's Accuracy

@@

Similarly to the previous analysis, the overall correlation of recognition and non-recognition revealed a significant but very low link between self-reported emotions and automatic \nolinebreak classifier's recognition (`r corr_sr_ar$full_result`). A confusion matrix was used to compare encoders’ subjective feeling with the emotion label recognized by the automatic classifier (Figure \ref{fig:confusionMatrix_sr_ar}).

```{r confusionMatrix_sr_ar, fig.height=6, fig.cap="\\label{fig:confusionMatrix_sr_ar}Confusion matrix of between the emotion self-reported as being characteristic of the elicitation with the emotion recognized by the automatic classifier."}
confusionMatrix_sr_ar_freq <- confusionMatrix_sr_ar$table %>% 
  as.data.frame() %>% 
  dplyr::rename(ar_emotion = Prediction, sr_emotion = Reference) %>% 
  dplyr::mutate(n_video = sum(Freq)) %>% 
  dplyr::mutate(prop_video = Freq/sum(Freq)) %>% 
  dplyr::group_by(sr_emotion) %>% 
  dplyr::mutate(n_sr = sum(Freq)) %>%
  dplyr::mutate(prop_sr = Freq/sum(Freq)) %>% 
  dplyr::ungroup()

@@

confusionMatrix_sr_ar_table %>%
  knitr::kable(
    "latex",
    caption = "\\label{table:confusionTable_sr_ar}Autonatic recognition accuracy metrics for each emotion.",
    booktabs = T,
    digits = 2,
    linesep = "",
    escape = FALSE,
    align = c('l',rep('c', 4))
  ) %>%
  kableExtra::kable_styling(
    latex_options = c("HOLD_position"),
    font_size = 8,
    full_width = TRUE
    ) %>%
   add_footnote("Note. \\textit{na.} values are produced when not enough data are available to compute accuracy indicators.", notation="none", escape = FALSE)
```

A comparable explanation involing the amount of undertermined video based on self-reports can be provided as the level of undetermined emotions are very high for the self reports.

## C. Comparison Between Human and Automatic Recognition
